---
title: "Contributing an atlas package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contributing an atlas package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
library(ggseg.extra)
```

ggseg atlases live in their own R packages, distributed through the
[ggsegverse r-universe](https://ggsegverse.r-universe.dev).
This gives users a single place to install from, whether your package is
hosted on GitHub, GitLab or elsewhere.

This article covers the last step: getting a finished atlas package listed on
r-universe. For building the package itself --- scaffolding it with
`setup_atlas_repo()`, writing the `data-raw/` build script, documenting the
atlas, and getting `R CMD check` to pass --- see
[Publishing an atlas](https://ggsegverse.github.io/ggseg.extra/articles/tutorial-publishing-an-atlas.html),
which walks through a complete example.

## What we expect of a package

1. The atlas works with ggseg, ggseg3d, or both.
2. It lives in its own R package.
3. That package is in a public version control repository.
4. It passes `R CMD check` without errors.

`setup_atlas_repo()` gives you a package that satisfies all four from the
start, including tests, GitHub Actions workflows and a pkgdown site. In
RStudio it is also available from the New Project dialog, under the ggseg
icon.

## Adding your package to r-universe

### 1. Fork the r-universe repository

Fork [ggsegverse/ggsegverse.r-universe.dev](https://github.com/ggsegverse/ggsegverse.r-universe.dev).

### 2. Add yourself to packages.json

`packages.json` lists every package in the ggsegverse r-universe. Add an
entry for yours:

```json
[
  {
    "package": "ggsegYeo2011",
    "url": "https://github.com/ggsegverse/ggsegYeo2011"
  },
  {
    "package": "yourpackage",
    "url": "https://github.com/yourusername/yourpackage"
  }
]
```

`package` must match the `Package` field of your DESCRIPTION exactly, and
`url` points at the repository. For a repository that is not on GitHub, or one
whose default branch is not the one to build, add `branch`:

```json
{
  "package": "yourpackage",
  "url": "https://gitlab.com/yourusername/yourpackage",
  "branch": "main"
}
```

### 3. Open a pull request

Include a sentence on what the atlas is and a link to the paper it comes from,
if there is one.

### 4. Review

We check that the package installs, that it passes `R CMD check`, and that the
atlas plots with ggseg and/or ggseg3d. Once merged, r-universe builds it
automatically and it appears at
[ggsegverse.r-universe.dev](https://ggsegverse.r-universe.dev), usually within
the hour.

## Once you are listed

Users install your atlas with:

```{r}
#| label: install-example
#| eval: false
install.packages(
  "yourpackage",
  repos = "https://ggsegverse.r-universe.dev"
)
```

r-universe rebuilds on every push to your repository, so updates need nothing
from you. If you move the repository, open another pull request to update its
`url` in `packages.json`.
