---
title: "Contributing an atlas package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contributing an atlas package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
library(ggseg.extra)
```

ggseg atlases live in their own R packages, distributed through the
[ggsegverse r-universe](https://ggsegverse.r-universe.dev).
This gives users a single place to install from, whether your package is
hosted on GitHub, GitLab or elsewhere.

This article covers the last step: getting a finished atlas package listed on
r-universe. For building the package itself --- scaffolding it with
`setup_atlas_repo()`, writing the `data-raw/` build script, documenting the
atlas, and getting `R CMD check` to pass --- see
[Publishing an atlas](https://ggsegverse.github.io/ggseg.extra/articles/tutorial-publishing-an-atlas.html),
which walks through a complete example.

## What we expect of a package

1. The atlas works with ggseg, ggseg3d, or both.
2. It lives in its own R package.
3. That package is in a public version control repository.
4. It passes `R CMD check` without errors.

`setup_atlas_repo()` gives you a package that satisfies all four from the
start, including tests, GitHub Actions workflows and a pkgdown site. In
RStudio it is also available from the New Project dialog, under the ggseg
icon.

## The core table

Every atlas carries a `core` table, one row per region. Four columns have a
fixed meaning across the whole ecosystem, and anything else is yours.

| Column    | Required | What it holds                                 |
| --------- | -------- | --------------------------------------------- |
| `label`   | yes      | the identifier, unique within the atlas       |
| `region`  | yes      | a short, readable key                         |
| `hemi`    | yes      | `"left"`, `"right"`, `"midline"`, or `NA`     |
| `display` | expected | the long-form name, for legends and axes      |

The pipelines fill all four in, so a generated atlas satisfies this without
you doing anything. The guidance below matters when you curate the result,
which most atlases need.

### `label` is the join key

`label` is what the palette and every geometry slot are keyed on, so it has
to be unique and it has to stay put. Rename it and the geometry decouples
from the metadata, which draws a region in the wrong colour, or not at all.
If you do need to re-key, `ggseg.formats::relabel_atlas()` rewrites all of
them together.

Keep the source parcellation's spelling here, even when it is ugly. A reader
tracing your atlas back to its publication needs that string.

### `region` is the key people type

`region` is the short form a user writes in their own results table, so it
should be the readable version of `label`: no hemisphere, no suffix that is
the same on every row. `lh_L_V1_ROI` becomes `v1 roi`; `Left-Thalamus`
becomes `thalamus`.

The pipelines derive it with `label_to_region()`, which strips the
hemisphere and normalises separators and case. It is a reasonable default,
not a rule --- it cannot know that `ROI` carries nothing or that `V1` should
keep its capitals. Where it gets your atlas wrong, declare a `region`
column in your lookup table and it is used exactly as you wrote it.

Two things to get right, because nothing can check them for you:

- **Do not leave the hemisphere in `region`.** If `region` is unique across
  the whole atlas when the parcellation is bilateral, the hemisphere is
  almost certainly still in there, and grouping by `region` then silently
  aggregates nothing.
- **Unless there genuinely is no pair.** Data-driven parcellations often
  number parcels globally, so a left parcel has no right counterpart to
  share a name with. Then `region` being unique is correct, and forcing a
  shared name would invent an anatomical claim the parcellation does not
  make. Say which situation yours is in, in the atlas's documentation.

### `hemi` is a closed vocabulary

`hemi` takes `"left"`, `"right"`, `"midline"` or `NA`, and nothing else is
accepted. It is the one column with no room for per-atlas judgement, because
plots filter and facet on it: a fifth spelling does not raise an error, it
drops those rows from the figure.

`NA` is for structures where a side does not apply. Use it rather than
inventing a value, and do not leave it `NA` for only part of an atlas ---
that is the case that bites, since `hemi == "left"` quietly loses rows.

A finer distinction than side belongs in a column of your own. The
cerebellar vermis is recorded as `"midline"`, for instance; if you need to
pick the vermis out, add a column that says so.

### `display` is for people

`display` is the long-form name, spelled to read well in a legend: `region`
`"bankssts"` has `display` `"banks of superior temporal sulcus"`. It is what
lets `region` stay terse without the figure becoming unreadable.

If you leave it out you get a message once per session telling you so. It
is not an error, because about twenty atlas packages were published before
the column existed, but a plot of your atlas has nothing but the machine key
to put in its legend.

### Everything else is yours

Add whatever columns classify your atlas usefully --- the bundled atlases
use `lobe`, `structure`, `group` and `label_short` --- and nothing
constrains their names or contents. They survive the atlas verbs and reach
the user's data frame, so a column like `network` or `lobe` is immediately
available to facet or colour by.

## Adding your package to r-universe

### 1. Fork the r-universe repository

Fork [ggsegverse/ggsegverse.r-universe.dev](https://github.com/ggsegverse/ggsegverse.r-universe.dev).

### 2. Add yourself to packages.json

`packages.json` lists every package in the ggsegverse r-universe. Add an
entry for yours:

```json
[
  {
    "package": "ggsegYeo2011",
    "url": "https://github.com/ggsegverse/ggsegYeo2011"
  },
  {
    "package": "yourpackage",
    "url": "https://github.com/yourusername/yourpackage"
  }
]
```

`package` must match the `Package` field of your DESCRIPTION exactly, and
`url` points at the repository. For a repository that is not on GitHub, or one
whose default branch is not the one to build, add `branch`:

```json
{
  "package": "yourpackage",
  "url": "https://gitlab.com/yourusername/yourpackage",
  "branch": "main"
}
```

### 3. Open a pull request

Include a sentence on what the atlas is and a link to the paper it comes from,
if there is one.

### 4. Review

We check that the package installs, that it passes `R CMD check`, and that the
atlas plots with ggseg and/or ggseg3d. Once merged, r-universe builds it
automatically and it appears at
[ggsegverse.r-universe.dev](https://ggsegverse.r-universe.dev), usually within
the hour.

## Once you are listed

Users install your atlas with:

```{r}
#| label: install-example
#| eval: false
install.packages(
  "yourpackage",
  repos = "https://ggsegverse.r-universe.dev"
)
```

r-universe rebuilds on every push to your repository, so updates need nothing
from you. If you move the repository, open another pull request to update its
`url` in `packages.json`.
