---
title: "Getting started with ggseg.extra"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ggseg.extra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
library(ggseg.extra)
```

ggseg.extra provides pipelines for creating brain atlas data sets compatible with the ggseg and ggseg3d plotting packages.

## The shortest path

From a FreeSurfer annotation to something you can plot:

```{r}
#| label: quickstart
#| eval: false
atlas <- create_cortical_from_annotation(
  input_annot = c("lh.aparc.annot", "rh.aparc.annot")
)

# Fewer vertices, and the voxel staircase rounded off
atlas <- atlas_polish(atlas, keep = 0.2)

plot(atlas)
```

That is the whole cortical pipeline. `sitrep()` reports whether your machine
has what the other pipelines need before you start one.

## Picking a creation function

Which pipeline you want depends on the format you have:

| Function | Input | Use case |
|----------|-------|----------|
| `create_cortical_from_annotation()` | FreeSurfer `.annot` files | Cortical parcellations (DK, DKT, Yeo networks) |
| `create_cortical_from_labels()` | Individual `.label` files | Custom region combinations |
| `create_cortical_from_gifti()` | GIFTI `.label.gii` files | Cortical parcellations in GIFTI format |
| `create_cortical_from_cifti()` | CIFTI `.dlabel.nii` files | HCP-style cortical parcellations |
| `create_cortical_from_neuromaps()` | Neuromaps `.func.gii` or `.nii` files | Brain maps and parcellations from neuromaps |
| `create_subcortical_from_volume()` | Volumetric segmentation | Subcortical structures (thalamus, amygdala) |
| `create_wholebrain_from_volume()` | Volumetric parcellation with colour table | Combined cortical + subcortical atlases |
| `create_cerebellar_from_gifti()` | GIFTI labels on the SUIT surface | Cerebellar parcellations |
| `create_cerebellar_from_annotation()` | FreeSurfer `.annot` on the SUIT surface | Cerebellar parcellations |
| `create_cerebellar_from_volume()` | Volumetric cerebellar segmentation | Cerebellar parcellations, with 3D meshes |
| `create_tract_from_tractography()` | Tractography files (`.trk`, `.tck`) | White matter tracts |
| `create_tract_from_volume()` | Volumetric tract labels | Tracts given as a segmentation rather than streamlines |

All functions produce a `ggseg_atlas` object that works with both ggseg (2D) and ggseg3d (3D).

### What's inside a ggseg_atlas

Every atlas contains:

- **Core metadata** --- region names, hemisphere labels, and the mapping between region names and annotation labels.
- **Colour palette** --- extracted from your input files or auto-generated.
- **3D data** --- vertex indices for cortical atlases, or meshes for subcortical and tract atlases.
- **2D geometry** (optional) --- sf polygon outlines for flat brain plots.

### How each pipeline works

The cortical pipeline projects inflated mesh triangles straight to 2D polygons,
so it takes seconds and needs no FreeSurfer installation. The subcortical and
tract pipelines instead build 3D meshes from a volume and get their 2D view by
slicing it. Each has its own `steps`, inputs and system requirements:
the [atlas workflows article](https://ggsegverse.github.io/ggseg.extra/articles/atlas-workflows.html)
walks through all five with diagrams.

### Post-processing

Raw atlases often contain regions you don't need (white matter, ventricles, unknown labels), views that don't show your structures well, and outlines that still follow the voxel grid.
The `atlas_region_*` and `atlas_view_*` verbs curate the first two and `atlas_polish()` shapes the third, all without rebuilding from scratch.
See `vignette("post-processing")` for the full toolkit and `?atlas-verbs` for the vocabulary.

## System requirements

Using pre-built atlases requires nothing beyond R.
Creating your own atlases may require additional R packages and system tools
depending on the pipeline:

- **Cortical** (annotation/label files) --- no system tools needed; uses the
  `freesurferformats` R package to read files and projects mesh to 2D
  directly.
- **Subcortical / whole-brain** ---
  [FreeSurfer](https://surfer.nmr.mgh.harvard.edu/); contours are traced from
  the projection itself, with no image round-trip.
- **Tract** --- no system tools; reads tractography files with R packages.

All heavier dependencies (`freesurfer`, `Rvcg`, `terra`, etc.) are in
Suggests and only loaded when needed.

Run `sitrep()` to check your setup, or see `vignette("system-setup")`
for details.

## Tutorials

Step-by-step tutorials that walk through complete atlas creation pipelines,
one per input format, live on the [package website](https://ggsegverse.github.io/ggseg.extra/)
under *Tutorials*. They are not shipped with the installed package because they
need FreeSurfer subject data and network access to build.

See also `vignette("pipeline-configuration")` for controlling verbosity, parallelism, and output directories.
