---
title: "Pipeline configuration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pipeline configuration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
library(ggseg.extra)
```

The atlas creation functions share common parameters that control pipeline behaviour.
You can set these explicitly in each function call, globally via R options, or through environment variables.

## Parameter hierarchy

Parameters resolve in this order:

1. **Explicit argument** --- value passed directly to the function
2. **R option** --- value from `options()`
3. **Environment variable** --- value from `Sys.getenv()`
4. **Default** --- built-in default value

This lets you set project-wide defaults while still overriding them for specific calls.

## Available options

| Parameter | R Option | Environment Variable | Default |
|-----------|----------|---------------------|---------|
| `verbose` | `ggseg.extra.verbose` | `GGSEG_EXTRA_VERBOSE` | `1` |
| `cleanup` | `ggseg.extra.cleanup` | `GGSEG_EXTRA_CLEANUP` | `TRUE` |
| `skip_existing` | `ggseg.extra.skip_existing` | `GGSEG_EXTRA_SKIP_EXISTING` | `TRUE` |
| `output_dir` | `ggseg.extra.output_dir` | `GGSEG_EXTRA_OUTPUT_DIR` | `tempdir()` |

`verbose` runs on a three-level scale rather than being a flag: `0` is silent,
`1` is progress, `2` adds FreeSurfer's own output. `TRUE` and `FALSE` are
accepted and mean `1` and `0`.

The `tolerance`, `smoothness`, `smooth_refinements` and `dilate` arguments are
gone, along with the `ggseg.extra.tolerance` and `ggseg.extra.smoothness`
options. Geometry shaping happens after the build: see `atlas_polish()` and
`vignette("post-processing")`. Passing one of the old arguments is an error.

## Setting options in R

Use `options()` to set defaults for your R session:

```{r}
#| label: set-options
#| eval: false
options(
  ggseg.extra.cleanup = FALSE
)

atlas <- create_cortical_from_annotation(
  input_annot = c("lh.aparc.annot", "rh.aparc.annot"),
  output_dir = "my_atlas"
)
```

### Verbosity

The `verbose` parameter controls progress messages during pipeline execution.

```{r}
#| label: verbose
#| eval: false
options(ggseg.extra.verbose = 0)

options(ggseg.extra.verbose = 2)

Sys.setenv(GGSEG_EXTRA_VERBOSE = "0")
```

### Cleanup

The `cleanup` parameter controls whether intermediate files are removed after pipeline completion.
Set it to `FALSE` to keep them for debugging:

```{r}
#| label: cleanup
#| eval: false
options(ggseg.extra.cleanup = FALSE)

atlas <- create_subcortical_from_volume(
  input_volume = "aseg.mgz",
  output_dir = "my_atlas_files"
)
```

### Skip existing

The `skip_existing` parameter lets you resume interrupted pipeline runs by reusing existing intermediate files:

```{r}
#| label: skip-existing
#| eval: false
options(ggseg.extra.skip_existing = FALSE)

options(ggseg.extra.skip_existing = TRUE)
```

### Output directory

Intermediate files — meshes, projections, traced contours — are written to a
folder named after the atlas under `output_dir`, and that folder is removed
again unless `cleanup = FALSE`. A build will not start if the folder already
holds files it did not write, because removing it would delete them: keep
source files somewhere other than `output_dir/atlas_name`, or pass
`cleanup = FALSE`. `output_dir` defaults to `tempdir()`, so a build leaves
nothing behind. Set it when you want the
intermediates to survive the session, which is what makes `skip_existing`
useful:

```{r}
#| label: output-dir
#| eval: false
options(ggseg.extra.output_dir = "~/atlas-builds")
```

### Geometry

Nothing about the geometry is configured here. The pipelines return raw
outlines and shaping is a separate step on the finished atlas, so there is
nothing to set before a build. See `vignette("post-processing")`.

## Environment variables

Environment variables are useful for CI pipelines, Docker containers, or settings that should persist across R sessions.

In `.Renviron`:

```
GGSEG_EXTRA_VERBOSE=0
GGSEG_EXTRA_CLEANUP=true
GGSEG_EXTRA_SKIP_EXISTING=true
```

In a shell:

```bash
export GGSEG_EXTRA_VERBOSE=0
R -e "ggseg.extra::create_cortical_from_annotation(...)"
```

In Docker:

```dockerfile
ENV GGSEG_EXTRA_VERBOSE=0
ENV GGSEG_EXTRA_CLEANUP=true
```

## Overriding defaults

Explicit arguments always win:

```{r}
#| label: override
#| eval: false
options(ggseg.extra.cleanup = TRUE)

atlas <- create_cortical_from_annotation(
  input_annot = c("lh.aparc.annot", "rh.aparc.annot"),
  cleanup = FALSE
)
```

## Recipes

### Development and debugging

```{r}
#| label: dev-options
#| eval: false
options(
  ggseg.extra.verbose = 2,
  ggseg.extra.cleanup = FALSE,
  ggseg.extra.skip_existing = FALSE
)
```

### Production and CI

```{r}
#| label: prod-options
#| eval: false
options(
  ggseg.extra.verbose = 0,
  ggseg.extra.cleanup = TRUE,
  ggseg.extra.skip_existing = TRUE
)
```

### Iterating on the geometry

Build once, then shape the returned atlas as many times as you like.
`atlas_polish()` does the usual pair — simplify, then smooth — against a vertex
budget:

```{r}
#| label: iterate
#| eval: false
annot_files <- c("lh.myatlas.annot", "rh.myatlas.annot")

atlas_raw <- create_cortical_from_annotation(
  input_annot = annot_files,
  output_dir = "atlas_workdir"
)

# High fidelity
atlas <- atlas_polish(atlas_raw, keep = 0.5)

# Compact, leaving the brain silhouette crisp
atlas <- atlas_polish(atlas_raw, keep = 0.05, exclude = context_pattern())
```

`context_pattern()` matches the grey silhouette the structures are read
against, which usually wants gentler treatment than they do. See
`vignette("post-processing")`.
