---
title: "Pipeline configuration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pipeline configuration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
library(ggseg.extra)
```

The atlas creation functions share common parameters that control pipeline behaviour.
You can set these explicitly in each function call, globally via R options, or through environment variables.

## Parameter hierarchy

Parameters resolve in this order:

1. **Explicit argument** --- value passed directly to the function
2. **R option** --- value from `options()`
3. **Environment variable** --- value from `Sys.getenv()`
4. **Default** --- built-in default value

This lets you set project-wide defaults while still overriding them for specific calls.

## Available options

| Parameter | R Option | Environment Variable | Default |
|-----------|----------|---------------------|---------|
| `verbose` | `ggseg.extra.verbose` | `GGSEG_EXTRA_VERBOSE` | `TRUE` |
| `cleanup` | `ggseg.extra.cleanup` | `GGSEG_EXTRA_CLEANUP` | `TRUE` |
| `skip_existing` | `ggseg.extra.skip_existing` | `GGSEG_EXTRA_SKIP_EXISTING` | `TRUE` |

The `tolerance`, `smoothness` and `smooth_refinements` parameters are deprecated.
sf simplification has moved out of atlas creation entirely — call `atlas_smooth()` on the returned atlas instead.

## Setting options in R

Use `options()` to set defaults for your R session:

```{r}
#| label: set-options
#| eval: false
options(
  ggseg.extra.cleanup = FALSE
)

atlas <- create_cortical_from_annotation(
  input_annot = c("lh.aparc.annot", "rh.aparc.annot"),
  output_dir = "my_atlas"
)
```

### Verbosity

The `verbose` parameter controls progress messages during pipeline execution.

```{r}
#| label: verbose
#| eval: false
options(ggseg.extra.verbose = FALSE)

Sys.setenv(GGSEG_EXTRA_VERBOSE = "false")
```

### Cleanup

The `cleanup` parameter controls whether intermediate files are removed after pipeline completion.
Set it to `FALSE` to keep them for debugging:

```{r}
#| label: cleanup
#| eval: false
options(ggseg.extra.cleanup = FALSE)

atlas <- create_subcortical_from_volume(
  input_volume = "aseg.mgz",
  output_dir = "my_atlas_files"
)
```

### Skip existing

The `skip_existing` parameter lets you resume interrupted pipeline runs by reusing existing intermediate files:

```{r}
#| label: skip-existing
#| eval: false
options(ggseg.extra.skip_existing = FALSE)

options(ggseg.extra.skip_existing = TRUE)
```

### Geometry parameters

sf geometry simplification is no longer applied during atlas creation.
Use `atlas_smooth()` on the returned atlas to control the trade-off between detail and file size.
See `vignette("post-processing")` for the full workflow.

## Environment variables

Environment variables are useful for CI pipelines, Docker containers, or settings that should persist across R sessions.

In `.Renviron`:

```
GGSEG_EXTRA_VERBOSE=false
GGSEG_EXTRA_CLEANUP=true
GGSEG_EXTRA_SKIP_EXISTING=true
```

In a shell:

```bash
export GGSEG_EXTRA_VERBOSE=false
R -e "ggseg.extra::create_cortical_from_annotation(...)"
```

In Docker:

```dockerfile
ENV GGSEG_EXTRA_VERBOSE=false
ENV GGSEG_EXTRA_CLEANUP=true
```

## Overriding defaults

Explicit arguments always win:

```{r}
#| label: override
#| eval: false
options(ggseg.extra.cleanup = TRUE)

atlas <- create_cortical_from_annotation(
  input_annot = c("lh.aparc.annot", "rh.aparc.annot"),
  cleanup = FALSE
)
```

## Recipes

### Development and debugging

```{r}
#| label: dev-options
#| eval: false
options(
  ggseg.extra.verbose = TRUE,
  ggseg.extra.cleanup = FALSE,
  ggseg.extra.skip_existing = FALSE
)
```

### Production and CI

```{r}
#| label: prod-options
#| eval: false
options(
  ggseg.extra.verbose = FALSE,
  ggseg.extra.cleanup = TRUE,
  ggseg.extra.skip_existing = TRUE
)
```

### Iterating on simplification level

`atlas_smooth(keep = ...)` is the single tuning knob for sf simplification.
Higher `keep` retains more vertices (more detail, larger file).
Try a few values without re-running the slow creation pipeline:

```{r}
#| label: iterate
#| eval: false
annot_files <- c("lh.myatlas.annot", "rh.myatlas.annot")

atlas_raw <- create_cortical_from_annotation(
  input_annot = annot_files,
  output_dir = "atlas_workdir"
)

# High fidelity
atlas <- atlas_raw |> atlas_smooth(keep = 0.5)

# Compact
atlas <- atlas_raw |> atlas_smooth(keep = 0.05, exclude = "cortex_")
```
