---
title: "Post-processing atlases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Post-processing atlases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
library(ggseg.extra)
```

Raw atlases come with everything: white matter, ventricles, unknown labels, cortex outlines, and views that may not show your structures well.
The `atlas_region_*` and `atlas_view_*` families let you curate an atlas without rebuilding from scratch.

Most of the functions in this vignette come from ggseg.formats and are re-exported by ggseg.extra for convenience.
The geometry adjustment functions (`atlas_smooth()`, `atlas_simplify()`) are native to ggseg.extra.

## Inspecting an atlas

Before changing anything, understand what you have:

```{r}
#| label: inspect
#| eval: false
library(ggseg.formats)

atlas_labels(atlas)

atlas_regions(atlas)

atlas_views(atlas)

atlas_sf(atlas)

atlas_meshes(atlas)
```

`atlas_labels()` returns the annotation-level identifiers.
`atlas_regions()` returns the display names stored in `$core`.
`atlas_views()` lists the available 2D views.
`atlas_sf()` returns the sf geometry used for 2D plotting.
`atlas_meshes()` returns the 3D mesh data for each region.

## Removing regions

The most common post-processing step is removing structures you don't need.
`atlas_region_remove()` matches against labels by default:

```{r}
#| label: remove
#| eval: false
atlas <- atlas |>
  atlas_region_remove("White-Matter") |>
  atlas_region_remove("WM-hypointensities") |>
  atlas_region_remove("-Ventricle") |>
  atlas_region_remove("-Vent$") |>
  atlas_region_remove("CSF")
```

Patterns are regular expressions, so `-Vent$` matches "3rd-Vent" and "4th-Vent" but not "Ventral-DC".

This removes the region from `$core`, `$palette`, and all geometry data.

## Keeping regions

The inverse operation --- keep only regions that match:

```{r}
#| label: keep
#| eval: false
atlas <- atlas |>
  atlas_region_keep("Thalamus|Caudate|Putamen|Pallidum")
```

Everything that doesn't match gets dropped.

## Context regions

Some regions work better as background outlines than as filled areas.
The cerebral cortex in a subcortical atlas is the classic example --- you want to see where it is for spatial reference, but you don't want it competing with the subcortical structures for colour.

`atlas_region_contextual()` keeps the geometry in `$data$sf` but removes the region from `$core`.
The region renders as an outline, not a filled polygon:

```{r}
#| label: contextual
#| eval: false
atlas <- atlas |>
  atlas_region_contextual("cortex", match_on = "label") |>
  atlas_region_contextual("unknown", match_on = "label") |>
  atlas_region_contextual("corpuscallosum", match_on = "label")
```

The `match_on` parameter controls whether patterns match against `label` (annotation identifiers) or `region` (display names).

## Labels versus regions

Every ggseg atlas carries two columns that describe the same parcels in different ways, and keeping them distinct matters.

The `label` column is the machine-readable identifier — the string that traces directly back to the source parcellation.
It should be unambiguous, stable across versions, and safe to use in programmatic joins: `"Left-Thalamus"`, `"lh_7Networks_3"`, `"ctx-lh-superiorfrontal"`.
When you merge external data (effect sizes, p-values, cortical thickness) onto the atlas, `label` is the column you match against.

The `region` column in `$core` is for humans.
It appears in plot legends and figure labels, so it should be readable, lowercase, and stripped of hemisphere prefixes and technical notation: `"thalamus"`, `"dorsal attention"`, `"superior frontal gyrus"`.
Hemispheres are already tracked in the `hemi` column — repeating them in `region` is redundant.

The pipeline generates both from the source annotation, but the defaults are rarely publication-ready.
`atlas_region_rename()` cleans up the `region` column without touching `label`:

```{r}
#| label: rename
#| eval: false
atlas <- atlas |>
  atlas_region_rename("Left-", "", match_on = "region") |>
  atlas_region_rename("Right-", "", match_on = "region") |>
  atlas_region_rename("-", " ", match_on = "region")
```

## Adding metadata columns

The `$core` data frame only requires three columns: `hemi`, `region`, and `label`.
Everything else is yours to add, and this is where atlases go from functional to genuinely useful.

Neuroscientists think in networks, lobes, and functional systems — not individual parcels.
Adding grouping columns lets users facet plots by system, color by network, or filter to structures they care about:

```{r}
#| label: core-add
#| eval: false
metadata <- data.frame(
  region = c("thalamus", "caudate", "putamen", "hippocampus", "amygdala"),
  structure = c("diencephalon", "basal ganglia", "basal ganglia",
                "limbic", "limbic")
)

atlas <- atlas |>
  atlas_core_add(metadata, by = "region")
```

This merges new columns into `$core` via a left join.

You can add as many columns as you need.
A Schaefer atlas might carry `network` and `parcel_order`.
A Brainnetome atlas might include `cytoarchitectonic_area` and `connectivity_profile`.
The atlas machinery ignores anything beyond the three required columns — extra columns are carried through plotting, subsetting, and serialization without interference.
The more semantic structure you encode, the more your users can do downstream without looking up region names in a separate table.

## View management

New subcortical and tract atlases have many slices that are selected by default to provide options of which slices give the best overall coverage of the atlas.
You will likely want to reduce the atlas to just views you see as necessary and provide the best general coverage.

### Keeping specific views

```{r}
#| label: view-keep
#| eval: false
atlas <- atlas |>
  atlas_view_keep("axial_3|axial_5|coronal_2|sagittal")
```

### Removing views

```{r}
#| label: view-remove
#| eval: false
atlas <- atlas |>
  atlas_view_remove("axial_1|axial_2")
```

### Reordering views

```{r}
#| label: view-reorder
#| eval: false
atlas <- atlas |>
  atlas_view_reorder(
    c("sagittal_left", "sagittal_right", "coronal_3", "axial_4")
  )
```

## Removing small regions from views

After filtering views, some regions may appear as tiny slivers that add clutter without information.
Remove them by minimum area:

```{r}
#| label: remove-small
#| eval: false
atlas <- atlas |>
  atlas_view_remove_region_small(
    min_area = 500,
    views = c("axial", "coronal")
  ) |>
  atlas_view_remove_region_small(min_area = 50)
```

The first call targets specific view types with a higher threshold.
The second call applies a lower threshold across all views.

## Gathering views

Raw atlas views often have large gaps between panels.
`atlas_view_gather()` repositions views into a compact layout:

```{r}
#| label: gather
#| eval: false
atlas <- atlas |>
  atlas_view_gather()
```

This is typically the last step before saving.

## Adjusting geometry after the fact

The atlas creation pipeline now returns raw, unsmoothed sf geometry.
Simplifying contours is a separate post-processing step you control after the atlas exists, so you can iterate on the result without re-running the (slow) creation pipeline.

`atlas_smooth()` is the single entry point for sf simplification.

### Smoothing rough contours

Region boundaries from volumetric or surface-based extraction tend to have staircase artefacts.
`atlas_smooth()` applies topology-preserving simplification via `rmapshaper::ms_simplify()`:

```{r}
#| label: smooth
#| eval: false
atlas <- atlas |>
  atlas_smooth(keep = 0.2)
```

The `keep` parameter is the proportion of vertices to retain (0–1).
Lower values produce simpler, smoother polygons.
Start around 0.2 and reduce further if the polygons still look noisy.

### Keeping the brain outline crisp

The cortex outline geometry doesn't usually want the same level of simplification as the labelled regions.
`exclude` matches labels that should be left untouched:

```{r}
#| label: smooth-exclude
#| eval: false
atlas <- atlas |>
  atlas_smooth(keep = 0.2, exclude = "cortex_")
```

Or use `labels` to simplify only matching labels and leave the rest alone.
Only one of `labels` or `exclude` may be supplied.

`atlas_smooth()` returns a modified `ggseg_atlas`, so you can inspect the result with `plot(atlas)` and adjust `keep` before committing.
The goal is an atlas that plots fast, looks neat, and shows regions correctly.

## Rebuilding the atlas

After modifying components directly (e.g., editing `$core` by hand), reconstruct the atlas to ensure consistency:

```{r}
#| label: rebuild
#| eval: false
atlas <- ggseg_atlas(
  atlas = atlas$atlas,
  type = atlas$type,
  palette = atlas$palette,
  core = modified_core,
  data = atlas$data
)
```

The constructor validates that core, palette, and geometry data are consistent.

## Putting it together

A typical post-processing pipeline for a subcortical atlas:

```{r}
#| label: full-pipeline
#| eval: false
atlas <- atlas_raw |>
  atlas_region_remove("White-Matter", match_on = "label") |>
  atlas_region_remove("-Ventricle", match_on = "label") |>
  atlas_region_remove("CSF", match_on = "label") |>
  atlas_region_contextual("Cortex", match_on = "label") |>
  atlas_view_keep("axial_3|axial_5|coronal_3|sagittal") |>
  atlas_view_remove_region_small(min_area = 100) |>
  atlas_smooth(keep = 0.2, exclude = "cortex_") |>
  atlas_view_gather()
```

Each step is a pure transformation --- pipe them together, inspect the result, adjust as needed.
