---
title: "System setup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{System setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
library(ggseg.extra)
```

Using pre-built atlases requires nothing beyond R.
Creating your own atlases may require external software depending on the
pipeline.

## What each pipeline needs

| Pipeline | R packages | System tools |
|----------|-----------|--------------|
| **Cortical** (annotation, labels) | `freesurferformats` | None |
| **Cortical** (neuromaps volume) | `freesurfer`, `neuromapr` | FreeSurfer |
| **Subcortical** | `freesurfer`, `Rvcg`, `terra` | FreeSurfer |
| **Whole-brain** | `freesurfer`, `RNifti`, `Rvcg`, `terra` | FreeSurfer |
| **Tract** | `RNifti`, `Rvcg` | None |
| **GIFTI / CIFTI** | `gifti` or `ciftiTools` | Connectome Workbench (CIFTI only) |

The cortical mesh-projection pipeline is the lightest --- it runs in seconds
with no system tools and minimal R dependencies. All heavier packages
(`freesurfer`, `Rvcg`, `terra`, etc.) are in Suggests and only
loaded when a pipeline actually needs them.

## FreeSurfer

[FreeSurfer](https://surfer.nmr.mgh.harvard.edu/fswiki/DownloadAndInstall)
is needed by the subcortical, whole-brain, and neuromaps volume pipelines.
The cortical annotation pipeline needs the `freesurferformats` R package
(not FreeSurfer itself) to read `.annot` and `.label` files.

On macOS, you also need:

- [XQuartz](https://www.xquartz.org/)
- [Xcode](https://developer.apple.com/xcode/) command line tools

Verify your installation:

```{r}
#| label: check-fs
#| eval: false
freesurfer::fs_dir()
```

## Parallel processing

ggseg.extra uses the [furrr](https://furrr.futureverse.org/) package for
parallel processing.
By default, processing runs sequentially.

To enable parallel processing, set up a `future` plan before running atlas
creation functions:

```{r}
#| label: parallel
#| eval: false
library(future)

plan(multicore, workers = 4)
```

**Prefer `multicore` where it is available.**
The pipelines hand each worker a whole volume, which under `multicore` costs
nothing: forked workers share the parent's memory copy-on-write.
A `multisession` worker is a fresh R process, so the volume is serialised to
it over a socket -- for a 256^3 aseg that is 67 MB per worker, which is more
work than the projection it was sent to compute.
Measured on an 8-core machine, the projection step runs about 3.4x faster
under `plan(multicore, workers = 4)` and slightly *slower* under
`plan(multisession, workers = 4)` than it does sequentially.

`multicore` is unavailable on Windows, and RStudio disables it by default.
There `multisession` is the only parallel option, and sequential is often the
better choice.

```{r}
#| label: parallel-windows
#| eval: false
plan(multisession, workers = 4)
```

To return to sequential processing:

```{r}
#| label: sequential
#| eval: false
plan(sequential)
```

## Progress bars

ggseg.extra uses [progressr](https://progressr.futureverse.org/) to report
progress during long-running operations.
Progress reporting is disabled by default.

To enable progress bars:

```{r}
#| label: progress
#| eval: false
library(progressr)

handlers("cli")
handlers(global = TRUE)
```

This gives you a cli-style progress bar that updates as each region or step
completes.
It works with both sequential and parallel execution.

## Checking your setup

Run the setup report to verify everything is in place:

```{r}
#| label: sitrep
#| eval: false
setup_sitrep()
```
