---
format: gfm
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 6,
  out.width = "100%",
  fig.retina = 3
)
```

# PKGNAME

<!-- badges: start -->
[![R-CMD-check](https://github.com/ggsegverse/PKGNAME/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ggsegverse/PKGNAME/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

ATLASNAME brain atlas for the ggseg ecosystem.

## Atlas Citation

> **TODO:** Add the original publication reference for this atlas here.

If you use this atlas in your work, please cite both the original atlas publication
and the ggseg package:

> Mowinckel AM, Vidal-Pineiro D (2020). "Visualization of Brain Statistics
> With R Packages ggseg and ggseg3d."
> *Advances in Methods and Practices in Psychological Science*, 3(4), 466-483.
> DOI: [10.1177/2515245920928009](https://doi.org/10.1177/2515245920928009)

## Installation

We recommend installing the ggseg-atlases through the ggsegverse
[r-universe](https://ggsegverse.r-universe.dev/#builds):

```{r}
#| label: install-r-universe
#| eval: false
options(repos = c(
  ggsegverse = "https://ggsegverse.r-universe.dev",
  CRAN = "https://cloud.r-project.org"
))

install.packages("PKGNAME")
```

You can install this package from [GitHub](https://github.com/) with:

```r
# install.packages("pak")
pak::pak("ggsegverse/PKGNAME")
```

## Usage

```{r}
#| label: plot-2d
#| eval: false
library(ggseg)
library(PKGNAME)

plot(ATLASNAME()) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 7)
  ) +
  guides(fill = guide_legend(ncol = 4))
```

## Code of Conduct

Please note that the PKGNAME project is released with a
[Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html).
By contributing to this project, you agree to abide by its terms.
