---
title: "Tutorial: Lookup tables and colours"
format: html
vignette: >
  %\VignetteIndexEntry{Tutorial: Lookup tables and colours}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

A volumetric parcellation is a NIfTI file full of integers.
Voxel 4 is `4`, and nothing in the file says that `4` means the left thalamus or that it should be drawn in teal.
That mapping lives in a separate text file — a colour lookup table, or LUT — and every volumetric pipeline in this package asks you for one.

If you got your parcellation from FreeSurfer, the LUT came with it.
If you got it from a paper's supplementary material, it probably didn't, and you have a spreadsheet of region names instead.
This tutorial covers both cases: reading a LUT, building one from nothing, and colouring it.

```{r}
#| label: load-libraries
library(ggseg.extra)
```

## What counts as a lookup table

A LUT is a plain data frame with six columns.
`idx` is the integer in the volume, `label` is the name, and `R`, `G`, `B` and `A` are the colour, each 0--255.
The optional columns `type`, `hemi` and `names` are covered [further down](#declaring-what-the-name-doesnt-say).

```{r}
#| label: minimal-lut
lut <- data.frame(
  idx = 0L,
  label = "Unknown",
  R = 0L, G = 0L, B = 0L, A = 0L
)

is_lut(lut)
```

`is_lut()` is the check every pipeline runs before it trusts your table, so it is worth calling yourself while you build one.
It is strict about names — a table with `index` and `name` columns is a perfectly good data frame and a useless LUT.

```{r}
#| label: not-a-lut
is_lut(data.frame(index = 1, name = "Thalamus"))
```

Row `idx = 0` is the background.
Every pipeline expects it, and several functions treat it specially, so leave it in place even when it looks like clutter.

## Adding regions

`lut_add()` appends a row and keeps the column types honest, which matters more than it sounds — a LUT with a double `idx` will not match the integers in your volume.

```{r}
#| label: lut-add
lut <- lut |>
  lut_add(idx = 1L, label = "Left-Thalamus", R = 0L, G = 118L, B = 14L) |>
  lut_add(idx = 2L, label = "Right-Thalamus", R = 0L, G = 118L, B = 14L) |>
  lut_add(idx = 3L, label = "Left-Putamen", R = 236L, G = 13L, B = 176L) |>
  lut_add(idx = 4L, label = "Right-Putamen", R = 236L, G = 13L, B = 176L)

lut
```

`A` defaults to `0`, which is what FreeSurfer writes and what the pipelines expect.
It is an opacity column that nothing in this ecosystem reads; it exists so the file format round-trips.

## Reading and writing

Real tables come from disk. `read_lut()` parses FreeSurfer's whitespace-separated format, and `write_lut()` writes it back:

```{r}
#| label: round-trip
path <- file.path(tempdir(), "demo_LUT.txt")
write_lut(lut, path)

cat(readLines(path), sep = "\n")
```

Reading it back gives you the same table, which is the property you want when a LUT is under version control alongside the atlas it describes:

```{r}
#| label: read-back
read_lut(path)
```

## Colours you didn't have to pick

Hand-picking 80 distinguishable colours is not a good use of an afternoon.
`lut_generate_colors()` spreads the palette around the colour circle for you, and — this is the part that makes it worth using — it gives both hemispheres of a structure the same colour, because `Left-Thalamus` and `Right-Thalamus` are one structure seen twice.

```{r}
#| label: generate-colors
blank <- data.frame(
  idx = 0:4,
  label = c(
    "Unknown",
    "Left-Thalamus", "Right-Thalamus",
    "Left-Putamen", "Right-Putamen"
  ),
  R = 0L, G = 0L, B = 0L, A = 0L
)

lut_generate_colors(blank)
```

Two structures, two colours, and `idx = 0` untouched.
Every other row is overwritten, so run this on a table whose palette you are willing to lose.

`chroma` and `luminance` control intensity and lightness.
If you ask for a chroma the display cannot render across that many hues, the function errors rather than quietly handing you two regions the same colour — lowering `chroma` is the usual fix.

```{r}
#| label: generate-colors-muted
lut_generate_colors(blank, chroma = 45)
```

### A trap with `by`

`by` names a column whose groups each get the *whole* colour circle instead of a slice of a shared one.
It is built for a whole-brain table split into `"cortical"` and `"subcortical"`, which become two atlases that never appear in the same plot — so a colour only has to be unique within each half.

The trap is that nothing stops you passing a column with one group per region.
Every group then starts at the same point on the circle, and you get one colour for the entire table:

```{r}
#| label: by-trap
grouped <- blank
grouped$region <- label_to_region(grouped$label)

lut_generate_colors(grouped, by = "region")
```

That is the documented behaviour working as designed, and it is still almost certainly not what you wanted.
Keep `by` for coarse splits — two or three groups, not forty.

## Hex colours for plotting

`get_lut()` adds the two columns plotting code actually wants: `color`, the hex string, and `roi`, the zero-padded index that ggseg3d matches on.

```{r}
#| label: get-lut
get_lut(lut_generate_colors(blank))
```

It takes a path as well as a data frame, so `get_lut("aseg_LUT.txt")` is a one-liner for inspecting a table on disk.

## Merging two tables

`lut_combine()` stacks tables, which comes up when you are bolting a custom subcortical parcellation onto a standard cortical one.

It warns about duplicate indices rather than resolving them, and the warning fires on the common case: both of your tables have a background row at `idx = 0`.

```{r}
#| label: combine
extra <- data.frame(
  idx = 0L, label = "Unknown",
  R = 0L, G = 0L, B = 0L, A = 0L
) |>
  lut_add(idx = 10L, label = "Left-Amygdala", R = 103L, G = 255L, B = 255L)

lut_combine(lut, extra)
```

Drop the duplicate background before combining and the warning goes away:

```{r}
#| label: combine-clean
lut_combine(lut, extra[extra$idx != 0L, ])
```

Take the warning seriously when it names an index other than `0`.
Two rows claiming `idx = 12` means one of your regions is about to silently become the other.

## From label to region name

Atlases carry two names per parcel.
`label` is the identifier from the source parcellation, stable and unambiguous; `region` is what a reader sees in a legend.

`label_to_region()` is the exact rule the pipelines use to derive one from the other — strip the hemisphere affix, turn separators into spaces, lower-case the result:

```{r}
#| label: label-to-region
label_to_region(c("Left-Thalamus", "Pu_Left", "Central_Lateral-Lateral_Posterior_Left"))
```

It is exported so that build scripts can reproduce the rule exactly.
If you are writing a metadata table keyed on `region`, key it on `label_to_region(label)` rather than on a `gsub()` you wrote by hand; that is how one region quietly joins to `NA`.

Check the output before trusting it.
The affixes it strips are a fixed set, and FreeSurfer's `ctx-lh-` convention is not among them:

```{r}
#| label: label-to-region-limits
label_to_region(c("ctx-lh-superiorfrontal", "ctx-rh-superiorfrontal"))
```

Both hemispheres kept their prefix, so one structure now looks like two different regions — and `region` is the column that pairs hemispheres.
When your labels use a convention it doesn't recognise, strip the affix yourself first.

## Handing the table to a pipeline

Every volumetric builder takes `input_lut`, and every one of them accepts either a path or a data frame:

```{r}
#| label: pipeline
#| eval: false
atlas <- create_subcortical_from_volume(
  input_volume = "my_parcellation.nii.gz",
  input_lut = lut,
  atlas_name = "myatlas",
  output_dir = "data-raw"
)
```

Passing the data frame is the better habit.
It means the table you inspected is the table the pipeline used, with no chance of editing one file and reading another.

## Declaring what the name doesn't say

The six columns are the minimum.
A lookup table can also carry columns that *declare* something about a label, and a pipeline that finds one uses it instead of working the answer out for itself.
Each answers one question, and all follow the same rule: a value the column cannot mean is an error before any work starts, and `NA` means "I'm not saying", which sends that one label back to the fallback.

### `type`: cortical, subcortical or cerebellar

For whole-brain volumes there is one more column to fill in.
`create_wholebrain_from_volume()` needs to know which labels are cortical, subcortical and cerebellar, and it will guess from surface coverage if you don't tell it — noisily, but it will guess.
The column takes exactly `"cortical"`, `"subcortical"` or `"cerebellar"`.
`"Cortical"` or `"cortex"` stops the build with the offending values named, rather than quietly dropping the label into the guess you were trying to avoid.
`lut_classify_anatomy()` works the column out properly by overlaying the volume on FreeSurfer's `aparc+aseg`:

```{r}
#| label: classify
#| eval: false
lut <- lut_classify_anatomy("my_parcellation.nii.gz", "my_LUT.txt")
write_lut(lut, "my_LUT.txt")
```

Run it once and commit the result.
It needs FreeSurfer installed, and there is no reason to pay for it on every build.

### `hemi`: which side a label is on

By default a label's hemisphere is read from its name: `Left-Thalamus`, `lh.something`, `region_L`, `ctx-rh-precuneus`.
That works until the names don't say — a parcellation numbered `Region_1` to `Region_268` has no side in its labels at all.
A `hemi` column says it outright:

```{r}
#| label: hemi-column
numbered <- data.frame(
  idx = 1:3,
  label = c("Region_1", "Region_2", "Region_3"),
  R = c(0L, 0L, 236L), G = c(118L, 118L, 13L), B = c(14L, 14L, 176L), A = 0L,
  hemi = c("left", "right", NA)
)

is_lut(numbered)
```

It takes `"left"` or `"right"` (`lh`/`rh` and `l`/`r` are read too), plus `"midline"` and, for the cerebellum, `"vermis"`.
The third row above declares nothing, so its hemisphere is still read from its name.

All three volumetric builders honour the column.
In `create_subcortical_from_volume()` and `create_cerebellar_from_volume()` it sets the region's hemisphere.
In `create_wholebrain_from_volume()` it also decides which cortical surface a parcel is kept on, so a parcel whose voxels spill across the midline doesn't become two regions.

### `names`: what a region is called in a legend

Every atlas the pipelines build carries three ways of referring to a region.
`label` is the identifier, unique and never edited.
`region` is the short key derived from it, the same on both sides of the brain.
`names` is the one meant to be read: the long-form name a plot shows.

Left alone, `names` is the region.
A `names` column in the table supplies something better, row by row:

```{r}
#| label: names-column
named <- data.frame(
  idx = c(10L, 49L, 28L),
  label = c("Left-Thalamus", "Right-Thalamus", "Left-VentralDC"),
  R = c(0L, 0L, 165L), G = c(118L, 118L, 42L), B = c(14L, 14L, 42L), A = 0L,
  names = c(NA, NA, "ventral diencephalon")
)

is_lut(named)
```

The two thalami declare nothing and are named `thalamus`, after their region.
`Left-VentralDC` would have been `ventraldc`.

Display names have spaces in them, which a lookup table file cannot hold in a single field, so `names` travels in a data frame and `write_lut()` leaves it out.
Keep it in the script that builds the table.

### Keeping the declarations in the file

Both columns survive a trip through a file, so the table you commit is the whole declaration.
`write_lut()` puts them after the colours, and when `hemi` is there it adds a comment line naming the fields, with `NA` wherever a row declares nothing:

```{r}
#| label: declared-round-trip
numbered$type <- c("cortical", "cortical", "subcortical")

path <- file.path(tempdir(), "numbered_LUT.txt")
write_lut(numbered, path)

cat(readLines(path), sep = "\n")
```

FreeSurfer skips the comment and reads only the first six fields of each line, so this is still a colour table as far as it is concerned.
`read_lut()` reads all of it back:

```{r}
#| label: declared-read-back
read_lut(path)
```

A table with `type` alone is written the older way — no comment line, `type` as a seventh field — so lookup table files you already have don't change.
Columns that declare nothing a pipeline reads, like the `roi` and `color` that `get_lut()` adds, stay out of the file.

## Where to go next

The LUT is the least glamorous file in an atlas build and the one that most often explains a broken result.
When a region comes out the wrong colour, missing, or merged with its neighbour, read the table first.

From here, [Post-processing atlases](post-processing.html) covers renaming and curating regions once the atlas exists, and the volumetric tutorials — [subcortical](tutorial-subcortical-atlas.html), [whole-brain](tutorial-wholebrain-atlas.html) and [cerebellar](tutorial-cerebellar-atlas.html) — put the table to work.
